Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed-particle hydrodynamics

### 2017-02-21 Manuscripts_GMD.tex
comments from dr. bursik

@@

\title{Plume-SPH 1.0: A 3D dust-gas volcano plume model based on Smoothed Particle Hydrodynamics}

%=======
\title{Plume-SPH 1.0: A 3D dust-gas volcano plume model based on smoothed particle hydrodynamics}
%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
% \Author[affil]{given_name}{surname}
\Author[1]{Zhixuan}{Cao}
\Author[1]{Abani}{Patra}
\Author[2]{Marcus}{Bursik}
\Author[3]{E. Bruce}{Pitman}
\Author[4]{Matthew}{Jones}
\affil[1]{Department of Mechanical and Aerospace Engineering, University at Buffalo, SUNY, New York, United State}
\affil[2]{Department of Geology, University at Buffalo, SUNY, New York, United State}
\affil[3]{Department of Material Design and Innovation, University at Buffalo, SUNY, New York, United State}
\affil[4]{Center for Computational Research,
University at Buffalo, SUNY, New York, United State}
%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
\runningtitle{Plume-SPH 1.0}
\runningauthor{Z. Cao et al}
%<<<<<<< HEAD

\correspondence{Abani Patra (abani@buffalo.com)}

%=======
%\correspondence{Abani Patra (abani.patra@gmail.com)}
%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}
%% These dates will be inserted by Copernicus Publications during the typesetting process.
\firstpage{1}
\maketitle
\begin{abstract}
%Volcanic ash, if it persists in the atmosphere, will pose great threat to aviation and communities. Therefore, accurate modeling and forecasts are crucial. 
VATDs (Volcanic ash transport and dispersion models) used for critical volcanic ash forecasts usually take outputs of eruption plume models as source terms. The accuracy of source terms is crucial for forecasts from such VATDs. All  existing 3D (three dimensional) plume models described in the literature use mesh based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantages of SPH in volcano plume modeling we adopt a relatively simpler model, a 3D dusty-gas dynamic model, targeted at capturing the salient features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. We also assume immediate dynamic and thermal dynamic equilibrium between air and erupted material and that the effect of wind is not significant.

Several newly developed techniques in SPH are adopted and extended to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture mixing at unresolved scales. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formalism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plume models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.
\end{abstract}
\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}

%The dynamics of explosive eruption clouds has been a central issue in volcanology  for a long time \citep{settle1978volcanic, wilson1978control}.
 Primary hazards associated with explosive grey volcanoes include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the upper atmosphere
% In addition, there are also other secondary hazards associated with volcanoes 
\citep{lockwood2013volcanoes}.
Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our focus here are  the risks that volcanic ash posed on aircrafts and communities. 
During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days.  VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide these source terms for VATDs and the forecast accuracy is critically dependent on these models. This paper reports on a new 3D (three dimensional) volcano plume model designed to exploit the advantages of mesh-free methods for 3D modeling of such plumes  that involve multi-phase free boundary flows.\\

@@

The development of 2D and 3D, time-dependent, and multiphase numerical models for volcanic plumes has provided new explanations for many features of explosive volcanism. One of the earliest of these is the 3D model PDAC (Pyroclastic Dispersion Analysis Code) \citep{neri2003multiparticle}  which is a non-equilibrium, multiphase, 3D compressible flow model. Conservation equations for each phase are solved separately with finite volume method.  A parallel computing version of PDAC was developed lately \citep{ongaro2007parallel}. Advanced numerical techniques, such as second order scheme and semi-explicit time stepping, was also adopted afterwards to improve the accuracy of PDAC \citep{carcano2013semi}. 
Another 3D model, SK-3D \citep{suzuki2005numerical} is a 3D time-dependent fluid dynamics model that attempts to reproduce the entrainment process of eruption clouds with relatively simpler physics but with high order numerical accuracy and high spatial resolution.
%which is applicable to time-dependent phenomena in actual volcanological simulations. 

@@

While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation in 2006 \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still pretty coarse compared with SK-3D and PDAC.
Besides adding to their special strengths (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC developers have moved to include the effect of microphysics into the model while ATHAM developer have extended its ability to modeling pyroclastic flow. Both are using finer and finer resolution. Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, was introduced based on three dimensional N phases Eulerian transportation equation. ASHEE is valid for low concentration and low Stokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adapting its unstructured Finite Volume solver.\\
To summarize, each 3D model has its own focus based on problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation and forecast skill (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques) .%are always a couple of conflict considerations in 3D plume simulations.\\
\subsection{Features of SPH}
All of these existing 3D models are mesh based Eulerian methods. To date there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcano plume simulation that we outline below. Among such Lagrangian methods, Smoothed Particle Hydrodynamics \citep{gingold1977smoothed,lucy1977numerical}, %even though relatively newer compared to well designed mesh based methods such as Finite Difference,
 based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far the most widely used meshfree scheme. Implementations of %SPH in volcanology  may be seen in \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.\\
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
and,
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These  will be enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcano plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})). So advection term will be treated exactly with SPH \citep{monaghan2005smoothed}.
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

@@

Open source availability and  the easy extension feature of SPH will facilitate developing of a more comprehensive community driven model.\\
\subsection{Our contributions}
As SPH is a relatively new method in computational fluid dynamics, implementations of SPH for compressible multiphase turbulent flows are few. Several issues endemic to classical SPH necessary for use in this domain are fixed in our implementation. The most popular applications of SPH (and their original motivating application) has been in simulating of free surface flow, such as breaking-waves and flood. Less attention was paid to velocity inlet and pressure outlet boundary conditions which are required in plume modeling. 
\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on time step is used to avoid the growth of numerical fluctuations near the pressure boundary. We impose velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{kumar2013parallel} is adapted to bypass the well-known tensile instability issues of classical SPH.\\
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing. The core solver of our model is parallelized by MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item What's more, imposing of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

The physics model of the plume is first presented in section \ref{sec:physics-model} and leads to a complete mathematical description of the volcanic plume (governing equations and boundary conditions). In section \ref{sec:SPH_method} we will briefly introduce the numerical tool, SPH method. Both the fundamental discretization formulation and techniques that are used to handle specific issues involved in plume modeling will be discussed. Verification and validation with numerical tests is presented in section \ref{sec:verification-validation}. In section \ref{sec:conclusion}, a discussion on future work is given following a brief summary.

\section{Physical Model} \label{sec:physics-model}
\subsection{Description of the model}
During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising kilometers to tens of kilometers until its heat is exhausted that buoyancy is lost. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. When plume rises up, relatively coarse particle might separate from main stream of the plume falling down onto the ground and probably be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor will condense to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor will further heat up entrained air and make the bulk density further dilute. The entrained vapor might also experience a similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation. 
All in all, the process of plume development is  essentially a multiphase turbulent mixing process coupled with heat transfer and many other microphysics and chemical reactions.

@@

Because of above assumptions, all other micro-physics processes (like phase change of  \texorpdfstring{H\textsubscript{2}O}, aggregation, decomposition, absorption of gas on the surface of solid, solution of gas into the liquid) and chemical process are not considered in this model. These ignored microphysics factors will come to play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not considered in this model yet. Introducing of wind effect in our model will need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually referred as weak plume. Our model also lacks the ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano at April 4th \citep{sigurdsson19841982, folch2016fplume}. We are focussed here on developing the SPH based methodoloy in the context of the more basic (and more critical) aspects of volcano plume and therefore devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods. Since our plan is an open source distribution of the tool we believe some of these enhancements will rapidly ensue with community participation.

\subsection{Governing equations}
Based above assumptions, the governing equation of our model is given as following, which is the same as governing equations of SK-3D \citep{suzuki2005numerical}.
\begin{align}
\dfrac{\partial \rho}{\partial t} + \nabla \cdot (\rho \textbf{v}) = 0 \label{eq:gov-cs-rho} \\
\dfrac{\partial \rho \xi}{\partial t} + \nabla \cdot (\rho \xi \textbf{v}) = 0 \label{eq:gov-cs-ks}\\
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot (\rho \textbf{v} \textbf{v} + p\textbf{I}) = \rho \textbf{g} \label{eq:gov-cs-v} \\
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot [(\rho E + p )\textbf{v}] = \rho \textbf{g} \cdot\textbf{v} \label{eq:gov-cs-e}
\end{align}
$\xi$ is the mass fraction of ejected material.
$E = e + K $ is total energy which is a summation of kinetic energy $K$ and internal energy $e$.
An additional equation is required to close the system. In this model, the equation for closing the system is the EOS (equation of state) for ideal gas.
p = (\gamma_m - 1)\rho e \label{eq:EOS}

>>>>

%%xx General points
%%xx 0 Some suggestions in the text itself are shown with "xx" -- so if you just search for "xx" you will find them
%%xx 1 Make sure tenses match throughout
%%xx 2 use \citet for citations in-line (text), e.g., "\citet{monaghanpaper} found that SPH is great...", not "Monaghan \citep{monaghanpaper} found that SPH is great..."
%%xx 3 Need a section on the atmospheric structure, and how atmosphere can be incorporated into model, especially NWP or radiosondes
%%xx 4 You often seem to forget to put a space between paragraphs.  LaTeX of course does not recognize a <return> as a paragraph delimiter, nor is it proper to end a normal paragraph with \\ or %
%%xx 5 You are inconsistent with capitalization in section headings.  Either always use an initial capital letter, or only use an initial capital letter on the first word, proper names or acronyms
%%xx 6 Figures are of generally poor quality -- use PDF and reproduce at larger scale.  
%%xx 7 Difficult to understand what figure of Pinatubo plume is supposed to be telling us about the model

@@

\title{Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed-particle hydrodynamics}

%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
% \Author[affil]{given_name}{surname}
\Author[1]{Zhixuan}{Cao}
\Author[1]{Abani}{Patra}
\Author[2]{Marcus}{Bursik}
\Author[3]{E. Bruce}{Pitman}
\Author[4]{Matthew}{Jones}
\affil[1]{Department of Mechanical and Aerospace Engineering, University at Buffalo, SUNY, New York, United State}
\affil[2]{Department of Geology, University at Buffalo, SUNY, New York, United State}
\affil[3]{Department of Material Design and Innovation, University at Buffalo, SUNY, New York, United State}
\affil[4]{Center for Computational Research,
University at Buffalo, SUNY, New York, United State}
%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
\runningtitle{Plume-SPH 1.0}
\runningauthor{Z. Cao et al}
%<<<<<<< HEAD

\correspondence{Abani Patra (abani@buffalo.edu)}

%=======
%\correspondence{Abani Patra (abani.patra@gmail.com)}
%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}
%% These dates will be inserted by Copernicus Publications during the typesetting process.
\firstpage{1}
\maketitle
\begin{abstract}
%Volcanic ash, if it persists in the atmosphere, will pose great threat to aviation and communities. Therefore, accurate modeling and forecasts are crucial. 
VATDs (Volcanic ash transport and dispersal models) used for volcanic ash forecasts sometimes take outputs of eruption plume models as source terms. The accuracy of source terms is crucial for forecasts from  VATDs. All  existing 3D (three dimensional) plume models described in the literature use mesh based methods. SPH (smoothed particle hydrodynamics), as a mesh free method, has several advantages over mesh based methods in modeling of multiphase free boundary flows. As an initial effort to exploit the feasibility and advantages of SPH in volcanic plume modeling, we adopt a relatively simple model, a 3D dusty-gas dynamic model, targeted at capturing the salient features of a volcanic plume. In this model, erupted material is assumed to be well-mixed, and represented by one phase while air is another phase. We also assume immediate dynamic and thermodynamic equilibrium between air and erupted material and that the effect of wind is not significant.

Several newly developed techniques in SPH are adopted and extended to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH methods. The $SPH-\varepsilon$ turbulence model is adopted to capture mixing at unresolved scales. Heat exchange due to turbulence is then calculated with a  Reynolds analogy. A corrected formalism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and in cross section with experimental results of JPUE (jet or plume that is ejected from a nozzle into a uniform environment). Next, the top height of the Pinatubo eruption of 15 June 1991, as simulated by our model, is consistent with both observations and existing 3D plume models. Profiles of several integrated variables are compared with those calculated in existing 3D plume models, and further verify our model. Analysis of the plume evolution process illustrates that this model is able to reproduce the basic physics of plume development. The comparison also implies that the turbulence model plays a significant role in 3D volcanic plume modeling.
\end{abstract}

\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}

%The dynamics of explosive eruption clouds has been a central issue in volcanology  for a long time \citep{settle1978volcanic, wilson1978control}.
 Primary hazards associated with explosive volcanic eruptions include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the atmosphere
% In addition, there are also other secondary hazards associated with volcanoes 
\citep{lockwood2013volcanoes}.
Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our focus here is  the hazards that volcanic ash poses to aircraft and communities.

During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days.  VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide these source terms for VATDs and the forecast accuracy is critically dependent on these models. This paper reports on a new 3D (three dimensional) volcano plume model designed to exploit the advantages of mesh-free methods for 3D modeling of such plumes  that involve multi-phase free boundary flows.

@@

The development of 2D and 3D, time-dependent, and multiphase numerical models for volcanic plumes has provided new explanations for many features of explosive volcanism. One of the earliest of these is the 3D model PDAC (Pyroclastic Dispersion Analysis Code) \citep{neri2003multiparticle}  which is a non-equilibrium, multiphase, 3D compressible flow model. Conservation equations for each phase are solved separately with finite volume method.  A parallel computing version of PDAC was developed lately \citep{ongaro2007parallel}. Advanced numerical techniques, such as second order scheme and semi-explicit time stepping, was also adopted afterwards to improve the accuracy of PDAC \citep{carcano2013semi}. 
Another 3D model, SK-3D \citep{suzuki2005numerical} is a 3D time-dependent fluid dynamics model that attempts to reproduce the entrainment process of eruption clouds with relatively simpler physics but with high order numerical accuracy and high spatial resolution.
%which is applicable to time-dependent phenomena in actual volcanological simulations.

@@

While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation in 2006 \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still pretty coarse compared with SK-3D and PDAC.
Besides adding to their special strengths (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC developers have moved to include the effect of microphysics into the model while ATHAM developer have extended its ability to modeling pyroclastic flow. Both are using finer and finer resolution. Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, was introduced based on three dimensional N phases Eulerian transportation equation. ASHEE is valid for low concentration and low Stokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adapting its unstructured finite volume solver.

To summarize, each 3D model has its own focus based on problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation and forecast skill (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques) .%are always a couple of conflict considerations in 3D plume simulations.\\
%
\subsection{Features of SPH}
All of these existing 3D models are mesh based Eulerian methods. To date there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcano plume simulation that we outline below. Among such Lagrangian methods, smoothed-particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical}, %even though relatively newer compared to well designed mesh based methods such as Finite Difference,
 based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.

Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
and,
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These  will be enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcano plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})). So advection term will be treated exactly with SPH \citep{monaghan2005smoothed}.
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

@@

Open source availability and  the easy extension features of SPH will facilitate development of a more comprehensive community driven model.

\subsection{Our contributions}
As SPH is a relatively new method in computational fluid dynamics, implementations of SPH for compressible multiphase turbulent flows are few. Several issues endemic to classical SPH necessary for use in this domain are fixed in our implementation. The most popular applications of SPH (and their original motivating application) has been in simulating of free surface flow, such as breaking-waves and flood. Less attention was paid to velocity inlet and pressure outlet boundary conditions which are required in plume modeling.

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on time step is used to avoid the growth of numerical fluctuations near the pressure boundary. We impose velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{kumar2013parallel} is adapted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing. The core solver of our model is parallelized by MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

The physical model of the plume is first presented in section \ref{sec:physics-model} and leads to a complete mathematical description of the volcanic plume (governing equations and boundary conditions). In section \ref{sec:SPH_method}, we  briefly introduce the numerical tool -- the SPH method. Both the fundamental discretization formulation and techniques that are used to handle specific issues involved in plume modeling are discussed. Verification and validation with numerical tests are presented in section \ref{sec:verification-validation}. In section \ref{sec:conclusion}, a discussion on future work is given following a brief summary.

\section{Physical Model} \label{sec:physics-model}
\subsection{Description of the model}
During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising kilometers to tens of kilometers until its heat is exhausted that buoyancy is lost. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. When plume rises up, relatively coarse particle might separate with main stream of the plume falling down onto the ground and probably be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor will condense to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor will further heat up entrained air and make the bulk density further dilute. The entrained vapor might also experience a similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation.

All in all, the process of plume development is  essentially a multiphase turbulent mixing process coupled with heat transfer and  other microphysical and chemical reactions.

@@

Because of the above assumptions, all other microphysical processes (such as the phase changes of \texorpdfstring{H\textsubscript{2}O}, aggregation, disaggregation, absorption of gas on the surface of solids, solution of gas into a liquid) and chemical processes are not considered in this model. These ignored microphysics factors will come to play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not considered in this model yet. Introducing of wind effect in our model will need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refereed as weak plume. Our model also lacks the ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano on 4 April 1985xx check year xx \citep{sigurdsson19841982, folch2016fplume}. We are focussed here on developing the SPH based methodoloy in the context of the more basic (and more critical) aspects of volcano plume and therefore devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods. Since our plan is an open source distribution of the tool we believe some of these enhancements will rapidly ensue with community participation.

\subsection{Governing equations}
Based above assumptions, the governing equation of our model is given as following, which is the same as governing equations of SK-3D \citep{suzuki2005numerical}.
\begin{align}
\dfrac{\partial \rho}{\partial t} + \nabla \cdot (\rho \textbf{v}) = 0 \label{eq:gov-cs-rho} \\
\dfrac{\partial \rho \xi}{\partial t} + \nabla \cdot (\rho \xi \textbf{v}) = 0 \label{eq:gov-cs-ks}\\
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot (\rho \textbf{v} \textbf{v} + p\textbf{I}) = \rho \textbf{g} \label{eq:gov-cs-v} \\
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot [(\rho E + p )\textbf{v}] = \rho \textbf{g} \cdot\textbf{v} \label{eq:gov-cs-e}
\end{align}
$\xi$ is the mass fraction of ejected material.
$E = e + K $ is total energy which is a summation of kinetic energy $K$ and internal energy $e$.
